Rare-event sampling applied to the simulation of extreme mechanical efforts exerted by a turbulent flow on a bluff body

### 2019-05-15 draft_p2.tex
Change title of figure density_streamlines
Minor modies following E's remarks

\begin{figure}
  \centering
  \includegraphics[width=.5\linewidth]{illustr_density_streamlines/illustr_density_streamlines.png}
  \caption{\label{fig:density+streamlines} Pressure field (in lattice units) and velocity streamlines at $t=t^{\star}$. In the near wake of the square, a (blocking) vortex blocks an intense vortex against the base of the obstacle. \EL{$f_d$ or $\tilde f_d$?}}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\linewidth]{illustr_AMS/figure_AMS}
  \caption{\label{fig:illustr_AMS} Selection-mutation procedure of the \ac{ams} \EL{ou TAMS?} algorithm to compute the probability that a real-valued observable $A(\mathbf{x}(t))$ reaches values larger than $Q$ throughout a trajectory of duration $T_a$.\EL{QUID of the score function mentioned in the text: is the observable?} The initial ensemble of blue trajectories (associated with the time evolution of the observable $A$) is first generated. These trajectories are independent and can be simulated in parallel. On the basis of the respective maxima $Q_1$, $Q_2$ and $Q_3$, the trajectory with the lowest maximum is discarded in the ensemble (dashed blue line). Among the two remaining trajectories, trajectory 3 is chosen randomly and copied until it reaches the value $Q_1$. It is then freely simulated from the branching point to the final time $T_a$. In case of deterministic dynamics, a small perturbation is introduced at the branching to separate the trajectories. This procedure is iterated $J$ times or until all trajectories exceed a fixed threshold $Q$.}
\end{figure}

% Introduction of AMS
%
The \acl{ams} algorithm \cite{cerou_adaptive_2007} inherits the basic ideas of {splitting algorithms} \cite{KahnHarris1951}, in which the sampling of a rare event is made easier by splitting the dynamical path of the system (in phase space) into 
%a sequence of events that can be sampled with a higher probability  
a sequence of ``short-cuts'' that lead more rapidly to the realization of this event \cite{glasserman_multilevel_1999,rolland_statistical_2015}. The related selection-mutation step of this algorithm is sketched in Fig.~\ref{fig:illustr_AMS}.
%
% As already mentioned in the introduction, the \ac{ams} has been successfully applied to a large body of problems for the past ten years. 
%
% Over the past ten years the \ac{ams} has been successfully applied to a large body of problems, including the computation of the dissociation time of biomolecules~\cite{teo_adaptive_2016} or the simulation of rare relaminarizations in a stochastic model of wall turbulence~\cite{rolland_extremely_2018}.
%
% Quick description of TAMS
%
We consider here a variant \EL{generalization?} of \ac{ams} called \ac{tams}, in which the branching procedure relies on a time-dependent \textit{score function} $\xi (\mathbf{x}(t),t)$. Each iteration therefore consists in discarding the trajectories with the lowest maxima of the score function $\xi (\mathbf{x}(t),t)$ during the time interval $0\leq t \leq T_a$. Discarded trajectories are re-sampled according to the remaining trajectories.
% As an illustration, a particular iteration of the \ac{tams} algorithm is sketched in figure~\ref{fig:illustr_AMS} .
Further details about the algorithm and corresponding mathematical results are provided in \cite{lestang_computing_2018} and references therein.
\EL{What is the difference between AMS and TAMS? why do we need TAMS rather than AMS, This is not mentioned in text.}

\subsubsection{Application of the \ac{ams} \EL{OR TAMS?} to a simple case: the \acl{ou} process}
\label{sec:ams_ou}

% Example: OU process
\begin{figure}
  \centering
  \includegraphics[width=.7\linewidth]{AMS_OU/AMS_OU.png}
    \caption{Efficiency of the \ac{tams} algorithm with respect to direct sampling in the case of an Ornstein-Uhlenbeck process \cite{lestang_computing_2018}. The orange line represents the evolution of the maximum obtained from re-sampled trajectories as a function of the computational cost $C_{TAMS}$. The solid blue line is the analytical solution for the return time of amplitude $a$. \EL{how many trajectories in the ensemble? REPLACE $f_d$ with $x$ in figure} }
  \label{fig:comparaison_temps_de_retour}
\end{figure}

@@

The \ac{tams} is applied  to a set of \EL{$N$} trajectories $\{x_n(t)\}_{0\leq t \leq T_a}$ with the objective to sample fluctuations of $x\geq a$ with $a$ being very large compared to typical values of $x$. The score function is here $x(t)$ and only one single trajectory is re-sampled at each iteration.
%
%Let $a_j$ be the maximum of the re-sampled trajectory at iteration $j$.
%
The computational cost of the algorithm after $J$ iterations is related to the simulation of the $N$ initial trajectories and the re-sampling of $J$ trajectories.
Fig.~\ref{fig:comparaison_temps_de_retour} compares the computational cost of the \ac{tams} algorithm with that of a direct sampling. 
In the latter, the typical computational cost is simply the return time $r(a)$.
Obviously, the successive re-samplings of the \ac{tams} algorithm lead very rapidly to trajectories exhibiting extreme fluctuations.
The related computational cost is many orders of magnitude lower than that of direct sampling.

@@

% Application to instantaneous drag
%
% Can the \ac{tams} algorithm achieve similar performance when applied to complex chaotic dynamics, such as turbulent flows around objects ?
% To answer this question, we consider again the two-dimensional test flow introduced in section~\ref{sec:test_flow}.
%
Our aim is now to sample flow evolution that exhibit extreme fluctuations of the drag $f_d$ acting on the square obstacle.
In contrast with the \acl{ou} process, the phase space is here higlhy-dimensional with intrinsically chaotic dynamics.
In that situation, the choice of the score function $\xi (\mathbf{x}(t),t)$ is no longer straightforward.
However, a natural simple choice for which we will opt in this study is certainly $\xi (\mathbf{x}(t),t) = f_d(t)$. 
\EL{say something about the perturbation applied at branching?}

@@

%pitfall
%
Fig.~\ref{fig:AMS_drag_trajectories} displays the ensemble of trajectories after many iterations of the \ac{tams} re-sampling procedure. 
%
Let us mention that for the sake of simplicity, the trajectory of the system is  here abusively associated with the time evolution of the observed drag.  \EL{useful to mention?}
%
Interestingly, it is obtained that all trajectories ultimately originate from one unique initial trajectory. 
%
This feature may be explained as follows.
Because the dynamics has a memory, it takes a certain time before a re-sampled trajectory separates from its parent. In our situation, this memory originates from the deterministic nature of fluid dynamics. 
Two trajectories (of the drag signal) starting from infinitesimally close points typically overlap over a timescale $\tau_L$ called the \textit{Lyapunov timescale}~\cite{lyapunovRef}. \EL{Freddy: Provide citation for Lyapunov timescale}
%
In our case, we observe in Fig.~\ref{fig:AMS_drag_trajectories} that the duration of extreme drag fluctuations is shorter than the Lyapunov timescale of the drag process.
Therefore, the re-sampling of a trajectory branched at $t=t^{\star}$ corresponding to the maximal drag amplitude cannot lead to larger values for $t^{\star} \leq t \leq t^{\star}+\tau_L$. For $t - t^{\star} >\tau_L$, the drag process has lost the memory of the drag fluctuations on which the re-sampling was based. Therefore, the probability of observing a new extreme fluctuation is also very low.
%

@@

\subsubsection{The \ac{gktl} algorithm}
\label{sec:gktl_description}
As with the \ac{tams} algorithm, the \ac{gktl} algorithm is based on the simulation of an ensemble of $N$ trajectories $\left\{\mathbf{x}_{n}(t)\right\}_{0\leq t \leq T_a}$ with $ n =1 \cdots N$ starting from independent random initial conditions.
%
Let us consider a real-valued observable of interest $A(\mathbf{x}(t))$ and introduce a cloning period $\tau$.
%
At time instants $t_{i}=i\tau$ with $i=1,~2,~...,~T_{a}/\tau$ ($T_{a}$ is a multiple of $\tau$) a weight $W_{n}^{i}$ is assigned to each trajectory. This weight is defined ($t_0=0$) by
%
\begin{equation}
W_{n}^{i}=\frac{e^{k\intop_{t_{i-1}}^{t_{i}}A(\mathbf{x}_{n}(t))dt}}{R_{i}}\quad \mbox{with the normalization factor} \quad R_{i}=\frac{1}{N}\sum_{n=1}^{N}e^{k\int_{t_{i-1}}^{t_{i}}A(\mathbf{x}_{n}(t))dt}
\label{eq:Weight}
\end{equation}
so that $\sum_{n=1}^N W_n^i = N$.
%
%
Therefore, the weights $\{W_{n}^{i}\}_{n=1\cdots N}$ determine (on average \EL{vague?}) how many copies of each trajectory are produced at time $t=t_i$. 
% For each t $\{\mathbf{x}_{n}\}_{0 \leq t \leq  t_i}$, a random number of copies of the trajectory are generated. This number is on average proportional to the weight $W_{n}^{i}$, such that the total number of trajectories produced at each event is equal to $N$.
%
For deterministic systems, a small perturbation is applied to the clones so that they can rapidly separate from their parents. 
%
% The parameter  is chosen by the user in order to control the strength of the selection.
%
The free parameter $k$ drives the amplitude of fluctuations targeted by the biased ensemble of trajectories.
A higher value of $k$ gives a higher weight to higher fluctuations and thus drives the sampling to more extreme events.

>>>>

\begin{figure}
  \centering
  \includegraphics[width=.5\linewidth]{illustr_density_streamlines/illustr_density_streamlines.png}
  \caption{\label{fig:density+streamlines} Pressure field (in lattice units) and velocity streamlines at $t=t^{\star}$. In the near wake of the square, a (blocking) vortex blocks an intense vortex against the base of the obstacle. \TL{changed to $\tilde f_d$.}}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\linewidth]{illustr_AMS/figure_AMS}
  \caption{\label{fig:illustr_AMS} Selection-mutation procedure of the \ac{ams} algorithm to compute the probability that a real-valued observable $A(\mathbf{x}(t))$ reaches values larger than $Q$ throughout a trajectory of duration $T_a$.\EL{QUID of the score function mentioned in the text: is the observable?} The initial ensemble of blue trajectories (associated with the time evolution of the observable $A$) is first generated. These trajectories are independent and can be simulated in parallel. On the basis of the respective maxima $Q_1$, $Q_2$ and $Q_3$, the trajectory with the lowest maximum is discarded in the ensemble (dashed blue line). Among the two remaining trajectories, trajectory 3 is chosen randomly and copied until it reaches the value $Q_1$. It is then freely simulated from the branching point to the final time $T_a$. In case of deterministic dynamics, a small perturbation is introduced at the branching to separate the trajectories. This procedure is iterated $J$ times or until all trajectories exceed a fixed threshold $Q$.}
\end{figure}

% Introduction of AMS
%
The \acl{ams} algorithm \cite{cerou_adaptive_2007} inherits the basic ideas of {splitting algorithms} \cite{KahnHarris1951}, in which the sampling of a rare event is made easier by splitting the dynamical path of the system (in phase space) into 
%a sequence of events that can be sampled with a higher probability  
a sequence of ``short-cuts'' that lead more rapidly to the realization of this event \cite{glasserman_multilevel_1999,rolland_statistical_2015}. The related selection-mutation step of this algorithm is sketched in Fig.~\ref{fig:illustr_AMS}.
%
% As already mentioned in the introduction, the \ac{ams} has been successfully applied to a large body of problems for the past ten years. 
%
% Over the past ten years the \ac{ams} has been successfully applied to a large body of problems, including the computation of the dissociation time of biomolecules~\cite{teo_adaptive_2016} or the simulation of rare relaminarizations in a stochastic model of wall turbulence~\cite{rolland_extremely_2018}.
%
% Quick description of TAMS
%
\TL{The selection of trajectories rely on a \textit{score function} $\xi (\mathbf{x}(t),t)$.}
Iterations of the algorithm consist in discarding the trajectories with the lowest maxima of the score function $\xi (\mathbf{x}(t),t)$ during the time interval $0\leq t \leq T_a$.
Discarded trajectories are re-sampled according to the remaining trajectories.
\TL{For a recent review of the \ac{ams}, its history and applications, see~\cite{cerou_2019}.}
\TL{Note that we use a specific form of the the \ac{ams} in which trajectories have a fixed duration $T_a$.
It can however be formulated in terms of the classical \ac{ams}~\cite{lestang_computing} and the following results are not limited to this particular variant.}

\subsubsection{Application of the \ac{ams} to a simple case: the \acl{ou} process}
\label{sec:ams_ou}

% Example: OU process
\begin{figure}
  \centering
  \includegraphics[width=.7\linewidth]{AMS_OU/AMS_OU.png}
    \caption{Efficiency of the \ac{tams} algorithm with respect to direct sampling in the case of an Ornstein-Uhlenbeck process \cite{lestang_computing_2018}. The orange line represents the evolution of the maximum obtained from re-sampled trajectories as a function of the computational cost $C_{TAMS}$. The solid blue line is the analytical solution for the return time of amplitude $a$. \TL{replaced $f_d$ with $x$ in figure} }
  \label{fig:comparaison_temps_de_retour}
\end{figure}

@@

\TL{The \ac{tams} is applied  to a set of $N=32$ trajectories $\{x_n(t)\}_{0\leq t \leq T_a}$ with $T_a=5$.Note that, for the process as define in~\eqref{eq:ou}, the correlation time is $\tau_c = 1$.}
Our objective to sample fluctuations of $x\geq a$ with $a$ being very large compared to typical values of $x$. The score function is here $x(t)$ and only one single trajectory is re-sampled at each iteration.
%
%Let $a_j$ be the maximum of the re-sampled trajectory at iteration $j$.
%
The computational cost of the algorithm after $J$ iterations is related to the simulation of the $N$ initial trajectories and the re-sampling of $J$ trajectories.
Fig.~\ref{fig:comparaison_temps_de_retour} compares the computational cost of the \ac{tams} algorithm with that of a direct sampling. 
In the latter, the typical computational cost is simply the return time $r(a)$.
Obviously, the successive re-samplings of the \ac{tams} algorithm lead very rapidly to trajectories exhibiting extreme fluctuations.
The related computational cost is many orders of magnitude lower than that of direct sampling.

@@

% Application to instantaneous drag
%
% Can the \ac{tams} algorithm achieve similar performance when applied to complex chaotic dynamics, such as turbulent flows around objects ?
% To answer this question, we consider again the two-dimensional test flow introduced in section~\ref{sec:test_flow}.
%
Our aim is now to sample flow evolution that exhibit extreme fluctuations of the drag $f_d$ acting on the square obstacle.
In contrast with the \acl{ou} process, the phase space is here higlhy-dimensional with intrinsically chaotic dynamics.
In that situation, the choice of the score function $\xi (\mathbf{x}(t),t)$ is no longer straightforward.
However, a natural simple choice for which we will opt in this study is certainly $\xi (\mathbf{x}(t),t) = f_d(t)$. 
\TL{Let us stress here that the flow dynamics are deterministic.
As a result, resampled trajectories do not differ from the parent trajectory their are resampled
from.
A workaround is to introduce a slight perturbation in the state of the flow at the branching point (see figure~\ref{fig:illustr_AMS}).
Because of the flow chaoticity, resampled trajectories will separate from their parent over the \textit{Lyapunov timescale} $\tau_L$.}\EL{Freddy: provide reference for Lyapunov timescale.}

@@

%pitfall
%
Fig.~\ref{fig:AMS_drag_trajectories} displays the ensemble of trajectories after many iterations of the \ac{tams} re-sampling procedure. 
%
Let us mention that for the sake of simplicity, the trajectory of the system is here abusively associated with the time evolution of the observed drag.  \EL{useful to mention?}\TL{Yes, good point!}
%
Interestingly, it is obtained that all trajectories ultimately originate from one unique initial trajectory. 
%
This feature may be explained as follows.
Because the dynamics has a memory, it takes a certain time $\tau_L$ before a re-sampled trajectory separates from its parent. In our situation, this memory originates from the deterministic nature of fluid dynamics. 
%
\TL{In our case, we observe in Fig.~\ref{fig:AMS_drag_trajectories} that the duration of extreme drag fluctuations is shorter than the Lyapunov timescale of the drag process.}
Therefore, the re-sampling of a trajectory branched at $t=t^{\star}$ corresponding to the maximal drag amplitude cannot lead to larger values for $t^{\star} \leq t \leq t^{\star}+\tau_L$.
For $t - t^{\star} >\tau_L$, the drag process has lost the memory of the drag fluctuations on which the re-sampling was based. Therefore, the probability of observing a new extreme fluctuation is also very low.
%

@@

\subsubsection{The \ac{gktl} algorithm}
\label{sec:gktl_description}
As with the \ac{tams} algorithm, the \ac{gktl} algorithm is based on the simulation of an ensemble of $N$ trajectories $\left\{\mathbf{x}_{n}(t)\right\}_{0\leq t \leq T_a}$ with $ n =1 \cdots N$ starting from independent random initial conditions.
%
Let us consider a real-valued observable of interest $A(\mathbf{x}(t))$ and introduce a cloning period $\tau$.
%
At time instants $t_{i}=i\tau$ with $i=1,~2,~...,~T_{a}/\tau$ ($T_{a}$ is a multiple of $\tau$) a weight $W_{n}^{i}$ is assigned to each trajectory. This weight is defined ($t_0=0$) by
%
\begin{equation}
W_{n}^{i}=\frac{e^{k\intop_{t_{i-1}}^{t_{i}}A(\mathbf{x}_{n}(t))dt}}{R_{i}}\quad \mbox{with the normalization factor} \quad R_{i}=\frac{1}{N}\sum_{n=1}^{N}e^{k\int_{t_{i-1}}^{t_{i}}A(\mathbf{x}_{n}(t))dt}
\label{eq:Weight}
\end{equation}
so that $\sum_{n=1}^N W_n^i = N$.
%
%
\TL{The weights $\{W_{n}^{i}\}_{n=1\cdots N}$ determine how many copies of each trajectory are produced at time $t=t_i$. See~\cite{brewer_efficient_2018, nemoto_population-dynamics_2016} for information on the practical implementation of the algorithm.}
% For each t $\{\mathbf{x}_{n}\}_{0 \leq t \leq  t_i}$, a random number of copies of the trajectory are generated. This number is on average proportional to the weight $W_{n}^{i}$, such that the total number of trajectories produced at each event is equal to $N$.
%
For deterministic systems, a small perturbation is applied to the clones so that they can rapidly separate from their parents. 
%
% The parameter  is chosen by the user in order to control the strength of the selection.
%
The free parameter $k$ drives the amplitude of fluctuations targeted by the biased ensemble of trajectories.
A higher value of $k$ gives a higher weight to higher fluctuations and thus drives the sampling to more extreme events.
